---
title: "STAT545_hw03_JLB"
output: github_document
---
```{r}
suppressPackageStartupMessages(library(tidyverse)) 
suppressPackageStartupMessages(library(gapminder))
```


** Task 1: Get the maximum and minimum of GDP per capita for all continents. **  

```{r}
gapminder %>%
  group_by(continent) %>% 
  summarize(minGdp = min(gdpPercap),
            maxGdp = max(gdpPercap)) %>% 
  knitr::kable()
  
```

** Task 2 Look at the spread of GDP per capita within the continents. **  
```{r}

```

** Task 3 Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean. **   
```{r}
#find the 5% trimmed mean of life expectancy. 
gapminder %>%
  group_by(year) %>%
  summarize(mean_lifeExp = mean(lifeExp, trim = .15)) %>% 
  knitr::kable()
  
```

** Task 4 How is life expectancy changing over time on different continents? **  
```{r}

lag(c(1,2,3,4))

#gapminder %>% 
 #group_by(continent) %>% 
 # mutate(gainLifeexp = lifeExp-lag(lifeExp)) %>% 
 # summarize(meanbycontinent = mean(gainLifeexp))
  #mutate(changeLEbyContinent = mean(gainLifeexp))
  
  
gapminder %>% 
 group_by(continent) %>% 
  mutate(meanLifeexp = mean(lifeExp))
  
  #summarize(meanbycontinent = mean(gainLifeexp))
  #mutate(changeLEbyContinent = mean(gainLifeexp))
  
  
#mutate(change_lifeExp =lifeExp - lag(lifeExp))) 
  
```

** Task 5 Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year. **    
```{r}

```

** Task 6 Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class. **   

```{r}

```

